---
title: "practicalML_courseProject"
author: "Nadia Stavisky"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
packages <- c("knitr", "caret", "pgmm","rpart", "gbm", "lubridate", "forecast", "e1071", "ggplot2")
# ipak function: install and load multiple R packages.
# check to see if packages are installed. Install them if they are not, then load them into the R session.
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

# usage
ipak(packages)
#set seed 12345
set.seed(12345)
```
# intorduction
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website [here](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

# Data used in the project:
- [The training data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv) 
- [The test data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

*[The source of the data for this project](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har)

```{r data_load, echo=FALSE}
test <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
train <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
```
Training data dimensious:
```{r}
dim(train)
```
We will use training data for building and assesing the different models.
Test data dimensious:
```{r}
dim(test)
```
Test data will be used with model that showed best performance for prediction.

# Data preparation:
- create subsets of train and test data, exclude columns that are mostly NA
```{r data_clean}
# remove variables that are mostly NA
allNA    <- sapply(train, function(x) mean(is.na(x))) > 0.95
train_data <- train[, allNA==FALSE]
# remove identification only variables (columns 1 to 7)
train_data <- train_data[,-c(1:7)]
dim(train_data)
```


# Project goal
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

```{r classe, }
str(train$classe)
classeh <- ggplot(train_data, aes(classe)) +
  geom_histogram(stat="count", color = "black", fill = "light blue") + 
  labs(title = "Predicting variable overview (train data set) \n 'classe' variable - is the exercise performnce manner type")
classeh
```

# Partition train data set on training and testing sets that would serve for model performance avaluation

```{r data_partition}
inTrain <- createDataPartition(y=train_data$classe, p=0.7, list = FALSE)
training <- train_data[inTrain,]
testing <- train_data[-inTrain,]
```

training data set dimensious:
```{r training_dim}
dim(training)
```
testing data set dimensious:
```{r testing_dim}
dim(testing)
```

# Building predicting models
Train Control ressampling - set method to "Cross validation"
Avaluation metrics - set avaluation metrics to "Accuracy"
```{r trainControl}
trCntr <- trainControl(method = "cv")
```

Fit different models 
- Random Forest;
```{r model_fit_rf}
modrf <- train(classe ~ ., method= "rf", trControl = trCntr, data = training)
```
- boosting with trees;
```{r model_fit_gbm}
modgbm <- train(classe ~ ., method= "gbm",  trControl = trCntr, data = training, verbose=FALSE)
```
- Linear discriminant analisis;
```{r model_fit_lda}
modlda <- train(classe ~ ., method= "lda", trControl = trCntr, data = training)
```
Avaluate models performance on training data set (train data).
Avaluation metrics - set avaluation metrics to "Accuracy"
```{r}
results <- resamples(list(lda=modlda, gbm=modgbm, rf=modrf))
summary(results)
```
Apply models on testing data set
```{r}
predictrf <- predict(modrf, newdata=testing)
confMatrf <- confusionMatrix(predictRF, testing$classe)
```
Avaluate models performance on testing data (train data).
Avaluation metrics - set avaluation metrics to "Accuracy"
```{r}

```
Predict the manner in which the exercise have been performed on initial test data
```{r}
predictTest <- predict(modrf, test)
predictTest
```
